Higher-Order unification for free!

### 2024-04-16 paper/main.tex
w-enforcement

The meta language of choice is expected to provide 
an implementation of \Ue that satisfies
the following properties:
%~\ref{prop:correct-ml} and~\ref{prop:complete-ml}.
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
\end{gather}

@@

\begin{lemma}[Compilation round trip]\label{prop:comprt} If
  \elpiIn{compile S T [] M [] _ [] _} then \elpiIn{decompile M T S}
\end{lemma}\noteD{Should also decompile links\dots}
% \begin{proof}[Proof sketch]
% Trivial since \mapStore is a bijection and 
% the terms are beta normal.
% some discussion about commit maybellam to be done later.
% \end{proof}

@@

\begin{lemma}[Properties of \Uo]\label{prop:uniffow} The following properties hold for \Uo:
\begin{gather}
  \unifcorrectX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}\\
  \unifcompleteX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}
\end{gather}
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{definition}[$\wellb/_\beta$]
$
\wellb/_\beta(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybeeta{} ~\cup~ \notllambda{})
$
\end{definition}

\begin{lemma}[Properties of \Uo]\label{prop:uniffowb} The following properties hold for \Uo
  where
  \begin{gather}
    \unifcorrectX{\Ee}{\Ue}{t}{\wellb/_\beta(\{t_1, t_2\})}\\
    \unifcompleteX{\Ee}{\Ue}{t}{\wellb/_\beta(\{t_1, t_2\})}
  \end{gather}
\end{lemma}

@@

\begin{lemma}[$\wellb/_\beta$-enforcement]\label{lem:w-enforcement-maybebeta}
  Given two terms $s_1$ and $s_2$ in \maybebeta, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
  then 
  $$
    \C{s_i}{t_i}{m_i}{l_i} ~\mathrm{for}~ i \in \{ 1,2 \} \Rightarrow
    t_1 \Ue t_2 \mapsto \sigma
  $$ 
\end{lemma}
\begin{proof}[Proof sketch]
  problematic terms are mapped to uva by comp,
  the problematic fapp node is gone.
  if F x = t then  F = lam x.t[]
  \todo{finish/maybe W enforcement is not needed here}
\end{proof}

\begin{theorem}[Fidelity in $\wellb/_\beta$] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb/_\beta(\foUnifPb)$
\end{theorem}
\begin{proof}[Proof sketch] thanks to \cref{lem:w-enforcement-maybebeta}
  \Ue is as powerful as \Uo in \maybebeta, as well as in \wellb by
  \cref{prop:uniffowb}.
\end{proof}

@@

@@

\begin{lemma}
  The introduction of \linketa guarantees \fullref{prop:w-enforcing}
  %\label{lemma:unif-wellb}
\end{lemma}

\begin{proof}[Proof sketch]
  By \cref{def:testmaybeeta}.\todo{shall we move the proof here?}
\end{proof}

@@

\begin{theorem}[Fidelity in \maybeeta]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb, if \noteD{riprendere sec 3.1}
  $\forall t, t\in\subterm{\foUnifPb} \Rightarrow t \notin \notllambda$ and if the 
  memory map is bijective, then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{This premise is fulfilled in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}
\todo{should we introduce W/eta as prev section W/beta?}

@@

@@

\begin{definition}[\progBetaLL]
  Let $\sigma$ be a substitution such that $\sigma t_1$ is a name $s$
  not occurring$s_1\dots s_n$. If $m = 1$, then $l$ is removed and \lhs is
  unified with $X_{s_1\ldots s_n~s}$.
  \todo{I changed this part, it is OK?}
  If $m > 1$, then $l$ is replaced by the link
  $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n~s}\appsep t_2\dots t_m}$
  (where $Y$ is a fresh variable of arity $n+1$) and link
  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n~x}}
  is added to \linkStore.

@@

\begin{proof}[Proof sketch]
  Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
  \progBetaRH{}, then it disappears from \linkStore and \elpiIn{progress}
  terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
  arguments. At each activation of \progBetaLL, $l$ is replaced by a new
  \linkbeta $l^1$ having $m-1$ arguments. At the $m^{th}$ iteration, the
  \linkbeta $l^m$ has no more arguments and is removed from \linkStore.
  Note that at the $m^{th}$ iteration, $m$ new \linketa have been added to
  \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
  terminates. Finally \progBetaFail also guarantees termination since 
  it makes \elpiIn{progress} immediately fails.

  % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  % \linkbeta disappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  % creates each time a new \linketa and replace the old \linkbeta with a new one.
  % This progression can however triggered at most $m$ times, since each new
  % \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  % the \linkbeta is removed from \linkStore which is now filled by $m$ new
  % \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  % terminates if \linkStore is made by only \linketa and therefore
  % \elpiIn{progress} still terminates.
  %\noteE[inline]{is it ok?}
  % \noteE[inline]{funziona. per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione). Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
\end{proof}

@@

\begin{theorem}[Fidelity with \linkbeta]
  The introduction of \linkbeta guarantees \fullref{prop:fidelity-recovery}
  if \Uo{} fails outside \llambda.
\end{theorem}
\todo{should we introduce W/llam as prev section W/beta and W/eta?}

>>>>

The meta language of choice is expected to provide 
an implementation of \Ue that satisfies
the following properties:
%~\ref{prop:correct-ml} and~\ref{prop:complete-ml}.
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
\end{gather}\todo{give a name and reuse it for \Uo in next sections}

@@

\begin{proposition}[Compilation round trip]\label{prop:comprt}
  % \elpiIn{compile F T [] M [] L [] S} then 
  % if \elpiIn{T} $\in$ S
  If $\C{s}{t}{m}{l}$, such that $l \in \linkStore$ and $m \in \mapStore$, if
  $\sigma = \{ A \mapsto t\}$ and $\mapping{X}{A}{n} \in \mapStore$
  then $\D{\sigma}{\mapStore}{\linkStore}{\rho}$ and
  $\rho X \Eo \rho s$.
\end{proposition}
\todo{explain no proof, llam link breaks}

% \noteD{Should also decompile links\dots}
% \begin{proof}[Proof sketch]
% Trivial since \mapStore is a bijection and 
% the terms are beta normal.
% some discussion about commit maybellam to be done later.
% \end{proof}

@@

\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow} The following properties hold for \Uo
  as implemented above: \todo{reuse statement from section 2.4}
\begin{gather}
  \unifcorrectX{\Eo}{\Uo}{t}{\wellb(\{t_1, t_2\})}\\
  \unifcompleteX{\Eo}{\Uo}{t}{\wellb(\{t_1, t_2\})}
\end{gather}
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{definition}[$\wellb_\beta$]
$
\wellb_\beta(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybeeta{} ~\cup~ \notllambda{})
$
\end{definition}

\begin{lemma}[Properties of \Uo in $\wellb_\beta$]\label{prop:uniffowb} The following properties hold for \Uo
  where \todo{reuse 2.4}
  \begin{gather}
    \unifcorrectX{\Ee}{\Ue}{t}{\wellb_\beta(\{t_1, t_2\})}\\
    \unifcompleteX{\Ee}{\Ue}{t}{\wellb_\beta(\{t_1, t_2\})}
  \end{gather}
\end{lemma}

@@

\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybebeta}
  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_\beta(s)$  then $\:\wellb(t)$.
\end{lemma}
% \begin{proof}[Proof sketch]
%   problematic terms are mapped to uva by comp,
%   the problematic fapp node is gone.
%   if F x = t then  F = lam x.t[]
%   \todo{finish/maybe W enforcement is not needed here}
% \end{proof}
\noindent
In other word the compiler eliminates all subterms that are in \maybebeta.

\begin{theorem}[Fidelity in $\wellb_\beta$] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb_\beta(\foUnifPb)$
\end{theorem}
\begin{proof}[Proof sketch] thanks to \cref{lem:w-enforcement-maybebeta}
  \Ue is as powerful as \Uo in \maybebeta, as well as in \wellb by
  \cref{prop:uniffowb}.
\end{proof}

@@

\begin{definition}[$\wellb_{\beta \eta}$]
  $
  \wellb_{\beta \eta}(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in \notllambda{}
  $
\end{definition}

\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybeeta}
  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_{\beta \eta}(s)$  then $\:\wellb(t)$.
\end{lemma}

\begin{proof}[Proof sketch]
  By \cref{lem:maybeeta}
\end{proof}

@@

@@

\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb, if \noteD{riprendere sec 3.1}
  $\wellb_{\beta \eta}(\foUnifPb)$ and if the 
  memory map is bijective, then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{This premise is fulfilled in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}
\todo{should we introduce W/eta as prev section W/beta?}

@@

% \begin{definition}[$\wellb_{\beta \eta \llambda}$]
%   $
%   \wellb_{\beta \eta}(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in \notllambda{}
%   $
% \end{definition}

\begin{lemma}[$\wellb$-enforcement]\label{lem:w-enforcement-maybellam}
  $\forall s, \C{s}{t}{m}{l}$, then $\:\wellb(t)$.
\end{lemma}

\begin{proof}[Proof sketch]
  By \cref{lem:w-enforcement-maybebeta,lem:w-enforcement-maybeeta} and the 
  rule above.
\end{proof}

@@

\begin{definition}[\progBetaLL]
  Let $\sigma$ be a substitution such that $\sigma t_1$ is a name $s$
  not occurring$s_1\dots s_n$. If $m = 1$, then $l$ is removed and \lhs is
  unified with $X_{s_1\ldots s_n~s}$.
  If $m > 1$, then $l$ is replaced by the link
  $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n~s}\appsep t_2\dots t_m}$
  (where $Y$ is a fresh variable of arity $n+1$) and link
  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n~x}}
  is added to \linkStore.

@@

\begin{proof}[Proof sketch]
  Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
  \progBetaRH{}, then it disappears from \linkStore and \elpiIn{progress}
  terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
  arguments. At each activation of \progBetaLL, $l$ is replaced by a new
  \linkbeta $l^1$ having $m-1$ arguments. At the $m^{th}$ iteration, the
  \linkbeta $l^m$ has no more arguments and is removed from \linkStore.
  Note that at the $m^{th}$ iteration, $m$ new \linketa have been added to
  \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
  terminates. Finally \progBetaFail also guarantees termination since 
  it makes \elpiIn{progress} immediately fails.
%
  % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  % \linkbeta disappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  % creates each time a new \linketa and replace the old \linkbeta with a new one.
  % This progression can however triggered at most $m$ times, since each new
  % \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  % the \linkbeta is removed from \linkStore which is now filled by $m$ new
  % \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  % terminates if \linkStore is made by only \linketa and therefore
  % \elpiIn{progress} still terminates.
  %\noteE[inline]{is it ok?}
  % \noteE[inline]{funziona. per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione). Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
\end{proof}

@@

\begin{theorem}[Fidelity in \Fo]
  The introduction of \linkbeta guarantees \fullref{prop:fidelity-recovery}
  if \Uo{} fails outside \llambda.
\end{theorem}
